The five axes of the Turtle: symmetry and asymmetry in NGC~6210

### 2020-10-03 ms.tex
Change all the slit letters in the text to lower-case

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figs/turtle-halo-slit-x}
  \caption{
    Deep images and spectra of the outer halo of NGC~6210, showing knotty structure.
    (a)~Composite mosaic of \oiii{} images,
    each with an exposure time of 1800~s,
    giving a combined field of view of \(3' \times 3'\) centered on the nebula.
    The position of slit position X is shown, for which a deep \Ha{} spectrum was obtained.
    Note that different sections of the mosaic are shown with different brightness normalizations,
    leading to abrupt jumps at the boundaries.
    The images were taken with an occulting bar that blocks out the bright core of the nebula,
    allowing the faint halo to be seen.
    (b)~\Ha{} spectrum from Slit~X, showing emission from two of the halo knots.
    The red dotted line shows the nebular nominal systemic velocity.
  }
  \label{fig:halo-knots}
\end{figure}

@@

All data was reduced using standard IRAF\footnote{IRAF was originally
  distributed by the National Optical Astronomy Observatory,
  \url{http://ast.noao.edu/data/software},
  and is currently available from \url{https://iraf-community.github.io}.}
routines.
Bias and cosmic-ray removal was followed by rectification and first-order wavelength 
calibration of the two-dimensional spectra
based on the comparison spectrum
of a Th/Ar lamp, yielding an accuracy in velocity units of \SI{\pm 1}{km.s^{-1}}.
Figure~\ref{fig:pv-array} shows example position-velocity arrays of (left-to-right)
\Ha{} \Wav{6563}, \oiii{} \Wav{5007}, and \nii{} \Wav{6583} for the 11 individual positions from the 2015 run.
The velocity scale is heliocentric 
and the scale along the slit is specified as arcsecond offsets from the central star.
% We derive a heliocentric
% systemic velocity, \(\vsys = \SI{40}{km.s^{-1}}\) by using the slit position F (epoch 2003) that passes through the central star and is not saturated. 
An additional slit (X, shown in Figure~\ref{fig:halo-knots}) was placed on the northern outskirts of the nebula 
in the outer halo of NGC~6210 where filaments had been clearly detected in some of our deep images.\footnote{For this spectrum only, \(3 \times 3\) pixel on-chip binning was used.}
% The observation was carried out on 18 September 2019.  The spectrum was obtained in order 87 (\Ha{} and \nii lines) with the 150 micron slit oriented at a position angle of 56 degrees.  The detector was again the Marconi 2048x2048 with a 3$\times$3 binning.  The spectral dispersion was 0.087 \AA/pix and 0.525\arcsec/pixel.
Most of the spectra presented here are publicly available from the \mbox{San Pedro Mártir} Kinematic Catalogue of Galactic Planetary Nebulae\footnote{
  \url{http://kincatpn.astrosen.unam.mx}
}
\citep{Lopez:2012a}.

@@

Along axis~B,
which is the apparent minor axis of the inner shell (\(\pm 5''\)),
the ellipse is distorted and the gradient is much more pronounced:
\(\pm \SI{11}{km.s^{-1}}\),
with the more negative velocities at the ENE end
(large blue circle symbols in lower panel of Fig.~\ref{fig:shell-velocity-axes}).
Velocity splitting of \(\pm \SI{9}{km.s^{-1}}\) is seen near both ends,
but it is not clear from the \oiii{} spectra if the ends are closed or open,
since none of the \oiii{} slits are aligned with this axis.
However, one of the \nii{} slits (slit~W) is indeed oriented close to axis~B and,
although the shell emits only weakly in \nii{},
the distorted ellipse is clearly closed at the ENE end
(large pink triangle symbol in Fig.~\ref{fig:shell-velocity-axes}).
The situation is not so clear at the WSW end
since any \nii{} emission from the shell is swamped by brighter emission from the knot complexes.
However, the evidence from \textit{HST} imaging suggests that the shell is closed in this direction also.

@@

Figure~\ref{fig:heii-shell-velocity-axes} shows the \heii{} velocity components projected on to the same two axes,
A and B,
which were found above from analysis of the \oiii{} shell kinematics.
The velocity and spatial scales are the same as in Figure~\ref{fig:shell-velocity-axes}
and the approximate loci of the \oiii{} shells are indicated by solid lines,
which allows the \oiii{} and \heii{} kinematics to be easily compared.
The maximum \heii{} velocity splitting of \(\pm \SI{13}{km.s^{-1}}\) is a little more than half as large as is seen in \oiii.
Along axis~A, the \heii{} emission is also more spatially compact than in \oiii{}.
Along axis~B, on the other hand, the \heii{} emission has a similar spatial extent to the inner \oiii{} shell,
even extending slightly beyond the inner shell on the WSW side
(for instance, the \heii{} component detected in slit~I\(^*\), as shown in the right panel of Fig.~\ref{fig:heii-shell-components}).
A gradient of \(\pm \SI{7.5}{km.s^{-1}}\) is seen along axis~B,
which has the same sense as the slightly larger gradient seen in \oiii{} along the same axis.

@@

As a check on our \Ha{}-derived densities, we also calculate densities from MES-SPM observations of the red \sii{} doublet for a single slit position (Slit~S of Fig.~2).
This passes through five distinct spatio-kinematic features in the inner nebula,
allowing us to measure the line ratio \(I(6716)/I(6731)\) for each
(unfortunately, \sii{} emission from the outer lobes and halo is undetectably weak).
Because the relative spectrograph efficiency between the doublet wavelengths is not well-determined,
we have calibrated the line ratios by forcing the average value over the entire slit to be \(I(6716)/I(6731) = 0.60\), in agreement with the spectrophotometry of \citet{Liu:2004a}.
Results are shown in Figure~\ref{fig:density-calibration},
calculated with PyNeb \citep{Luridiana:2015a} using atomic data from \citet{Podobedova:2009a} and \citet{Tayal:2010a}.
It can be seen that there is a reasonable agreement between the two density diagnostics,
which justifies our decision not to use any filling factor correction in equation~\eqref{eq:sha-density}.
The need for a filling factor is obviated by carefully calculating the line-of-sight depth of each emission component separately.

>>>>

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figs/turtle-halo-slit-x}
  \caption{
    Deep images and spectra of the outer halo of NGC~6210, showing knotty structure.
    (a)~Composite mosaic of \oiii{} images,
    each with an exposure time of 1800~s,
    giving a combined field of view of \(3' \times 3'\) centered on the nebula.
    The position of slit~x is shown, for which a deep \Ha{} spectrum was obtained.
    Note that different sections of the mosaic are shown with different brightness normalizations,
    leading to abrupt jumps at the boundaries.
    The images were taken with an occulting bar that blocks out the bright core of the nebula,
    allowing the faint halo to be seen.
    (b)~\Ha{} spectrum from slit~x, showing emission from two of the halo knots.
    The red dotted line shows the nebular nominal systemic velocity.
  }
  \label{fig:halo-knots}
\end{figure}

@@

All data was reduced using standard IRAF\footnote{IRAF was originally
  distributed by the National Optical Astronomy Observatory,
  \url{http://ast.noao.edu/data/software},
  and is currently available from \url{https://iraf-community.github.io}.}
routines.
Bias and cosmic-ray removal was followed by rectification and first-order wavelength 
calibration of the two-dimensional spectra
based on the comparison spectrum
of a Th/Ar lamp, yielding an accuracy in velocity units of \SI{\pm 1}{km.s^{-1}}.
Figure~\ref{fig:pv-array} shows example position-velocity arrays of (left-to-right)
\Ha{} \Wav{6563}, \oiii{} \Wav{5007}, and \nii{} \Wav{6583} for the 11 individual positions from the 2015 run.
The velocity scale is heliocentric 
and the scale along the slit is specified as arcsecond offsets from the central star.
% We derive a heliocentric
% systemic velocity, \(\vsys = \SI{40}{km.s^{-1}}\) by using the slit position F (epoch 2003) that passes through the central star and is not saturated. 
An additional slit (x, shown in Figure~\ref{fig:halo-knots}) was placed on the northern outskirts of the nebula 
in the outer halo of NGC~6210 where filaments had been clearly detected in some of our deep images.\footnote{For this spectrum only, \(3 \times 3\) pixel on-chip binning was used.}
% The observation was carried out on 18 September 2019.  The spectrum was obtained in order 87 (\Ha{} and \nii lines) with the 150 micron slit oriented at a position angle of 56 degrees.  The detector was again the Marconi 2048x2048 with a 3$\times$3 binning.  The spectral dispersion was 0.087 \AA/pix and 0.525\arcsec/pixel.
Most of the spectra presented here are publicly available from the \mbox{San Pedro Mártir} Kinematic Catalogue of Galactic Planetary Nebulae\footnote{
  \url{http://kincatpn.astrosen.unam.mx}
}
\citep{Lopez:2012a}.

@@

Along axis~B,
which is the apparent minor axis of the inner shell (\(\pm 5''\)),
the ellipse is distorted and the gradient is much more pronounced:
\(\pm \SI{11}{km.s^{-1}}\),
with the more negative velocities at the ENE end
(large blue circle symbols in lower panel of Fig.~\ref{fig:shell-velocity-axes}).
Velocity splitting of \(\pm \SI{9}{km.s^{-1}}\) is seen near both ends,
but it is not clear from the \oiii{} spectra if the ends are closed or open,
since none of the \oiii{} slits are aligned with this axis.
However, one of the \nii{} slits (slit~w) is indeed oriented close to axis~B and,
although the shell emits only weakly in \nii{},
the distorted ellipse is clearly closed at the ENE end
(large pink triangle symbol in Fig.~\ref{fig:shell-velocity-axes}).
The situation is not so clear at the WSW end
since any \nii{} emission from the shell is swamped by brighter emission from the knot complexes.
However, the evidence from \textit{HST} imaging suggests that the shell is closed in this direction also.

@@

Figure~\ref{fig:heii-shell-velocity-axes} shows the \heii{} velocity components projected on to the same two axes,
A and B,
which were found above from analysis of the \oiii{} shell kinematics.
The velocity and spatial scales are the same as in Figure~\ref{fig:shell-velocity-axes}
and the approximate loci of the \oiii{} shells are indicated by solid lines,
which allows the \oiii{} and \heii{} kinematics to be easily compared.
The maximum \heii{} velocity splitting of \(\pm \SI{13}{km.s^{-1}}\) is a little more than half as large as is seen in \oiii.
Along axis~A, the \heii{} emission is also more spatially compact than in \oiii{}.
Along axis~B, on the other hand, the \heii{} emission has a similar spatial extent to the inner \oiii{} shell,
even extending slightly beyond the inner shell on the WSW side
(for instance, the \heii{} component detected in slit~i\(^*\), as shown in the right panel of Fig.~\ref{fig:heii-shell-components}).
A gradient of \(\pm \SI{7.5}{km.s^{-1}}\) is seen along axis~B,
which has the same sense as the slightly larger gradient seen in \oiii{} along the same axis.

@@

As a check on our \Ha{}-derived densities, we also calculate densities from MES-SPM observations of the red \sii{} doublet for a single slit position (slit~s of Figure~\ref{fig:slit-positions}).
This passes through five distinct spatio-kinematic features in the inner nebula,
allowing us to measure the line ratio \(I(6716)/I(6731)\) for each
(unfortunately, \sii{} emission from the outer lobes and halo is undetectably weak).
Because the relative spectrograph efficiency between the doublet wavelengths is not well-determined,
we have calibrated the line ratios by forcing the average value over the entire slit to be \(I(6716)/I(6731) = 0.60\), in agreement with the spectrophotometry of \citet{Liu:2004a}.
Results are shown in Figure~\ref{fig:density-calibration},
calculated with PyNeb \citep{Luridiana:2015a} using atomic data from \citet{Podobedova:2009a} and \citet{Tayal:2010a}.
It can be seen that there is a reasonable agreement between the two density diagnostics,
which justifies our decision not to use any filling factor correction in equation~\eqref{eq:sha-density}.
The need for a filling factor is obviated by carefully calculating the line-of-sight depth of each emission component separately.
